\pagestyle{plain}
\pagenumbering{roman}
\setcounter{page}{3}

\section*{Acknowledgements}
In bringing this project to a conclusion I would first like to express my profound gratitude to my three excellent supervisors David Wright, Anders Romundset and Ingrid Fuglestvedt. David has by some mysterious ability allowed me both ample creative room to aimlessly explore interesting albeit not always as fruitful ideas, while at the same time making sure to keep me firmly on track. David's effectiveness, thoroughness, extensive knowledge of the discipline, and help with maneuvering small and large peculiarities of the academic world has formed a thoroughly dependable foundation on which to undertake this project. Anders has been crucial for all my dealings with geology through the project, and was instrumental for productively relating relative sea-level change to the archaeological record. Anders has shared generously and pedagogically of his time and knowledge when reading texts, during two rounds of fieldwork in the Oslo region and when showing me the ropes of the subsequent laboratory work during my visit to the Geological Survey in Trondheim. Finally, Ingrid has read my texts with great enthusiasm and encouragement, and her overall curiosity and enthusiasm for seemingly all aspects of archaeology has been very motivational, whether shared through supervision or informal talks in the break-room.

I would also like to thank the different research groups that I have been be a part of through the project. This includes Prehcoast and Archaeology by Proxy at the Museum of Cultural History, and the Materialities group at the Department of Archaeology, Conservation and History. Through these I have been able to present and discuss my work with a great range of archaeologists, students of archaeology, and non-archaeologists, both formally and informally -- offering both welcome delves and escapes from the depths of Mesolithic and computational archaeology.

Steinar Solheim, who is also co-author of the final paper of this thesis, has provided valuable feedback and continous encouragement both in the time before and throughout the project. Per Persson has also shared extensively from his own critical engagement with shoreline dating and his wide knowledge of the Scandianavian Stone Age more generally. I am also very grateful to Daniel Groß who evaluated my work for my midway assessment and had many valuable perspectives. As my MSc in Computational Archaeology from UCL in 2018/2019 was also instrumental to this project, I wish to thank Mark Lake and Andrew Bevan for having developed and coordinated a degree programme that still inspires four years on.

All the people at the department and Blindernveien 11 also deserve great thanks for making it a stimulating place to have spent most of my days over the last three years. Knut Ivar Austvoll deserves a special thanks for borrowing me extra computer power towards the end of the project, but especially for many relevant and irrelevant discussions. My fellow PhD students, including those at the Museum of Cultural History, were also instrumental to making this a thoroughly enjoyable experience. Especially Hallvard Bruvoll, who started one month before me and also had a quantitative slant to his project was an excellent canary ahead of me in the coal mine.

Irmelin Axelsen, Weronika Patrycja Polanska and Steinar Solheim read and had valuable feedback in the finishing stages of writing the introductory text. Finally, the disciplinary and administrative support of the PhD-programme at the department has been excellent, and especially the courses offered through DialPast have been invaluable.

\bigskip
\noindent
Oslo, September 2023
\newpage

\section*{Summary}
Through a series of case-studies that employ Mesolithic data from the Skagerrak coast of south-eastern Norway, this thesis is concerned with contributing to getting a handle on the vast amounts of data that have been and continue to be generated by Mesolithic archaeology in Norway. Fundamental questions concerning the quantity, composition and chronological control we have of the material available to us dictates not only our understanding of the period, but also determines what questions we can hope to answer. This follows from the resolution and quality of the archaeological record, and the varying spatial and temporal scales that different behavioural and societal dynamics operate over. The degree to which the characteristics of our data matches those necessary to illuminate these dynamics thus informs the kind of archaeology we can reliably hope to do.   

Given that this data is neither recorded or can be approached without a predefined understanding of what dimensions of it are of interest, the material was explored using a series of heuristic models that are frequently drawn on in the literature to understand prehistoric hunter-gatherer societies and the coastal Mesolithic of Northern Europe. These pertain especially to the dating of the archaeological material, patterns of land-use and mobility, as well as general trends of demographic development. 

The thesis consists of four papers and an introductory text. The introductory text establishes the context and motivation for undertaking the studies, relates them to each other, and presents further avenues along which the results can be explored, extended, and potentially be substantiated further. With an aim to make these efforts transparent, reproducible and extendable, the entirety of the project has been undertaken within a framework of open science. This means that all text, figures, data and code is made freely available for anyone to access and scrutinise. Along the same lines, and in part following from the quantitative nature of the derived results, all substantive inferences are instantiated as explicit causal hypotheses in the introductory text. The purpose of this is to promote clear routes for critical engagement and further research efforts.  

Some central results from the papers of the thesis follow first from a quantitative assessment of the relationship between coastal Stone Age sites and the contemporaneous shoreline in the region. This largely verified the previously suggested tendency of the sites to have been located close to the shoreline when they were occupied. Following from the quantitative nature of this assessment, this resulted in the development of a new method for shoreline dating Stone Age sites in the region, based on their altitude relative to the present-day sea-level. This method for shoreline dating has been developed as freely available package for the R programming language through the second paper of the thesis, making the method readily accessible for anyone to apply and assess. 

The third paper of the thesis focused on approaches for analysing the composition of a larger number of lithic assemblages by structuring the analysis through the use of multivariate statistics. First, this was done to assess the temporal development in the occurrence of artefact types that are in established use in Norwegian archaeology, with the findings largely verifying the current general understanding of these developments. More novel insights follow from the second line of investigation which explored dimensions of the assemblages that have frequently been drawn on outside the Fennoscandian setting to map variation in land-use and mobility patterns associated with the formation of lithic assemblages. Specifically, the volumetric density of lithics on the sites and the proportion of these that have been subjected to secondary modification were found to be negatively correlated. Furthermore, this was also found to follow a temporal development with an increase in the density of lithics and decrease in secondarily worked lithics over time. Drawing on the substantive implications that this relationship has been ascribed in other contexts, the findings could indicate a corresponding overall decrease in occupational duration at the sites through the period. While the relevance of this framework for the Fennoscandian Mesolithic is in need of further substantiation and evaluation, these developments match those previously suggested in the literature concerned with the Mesolithic of south-eastern Norway, giving some additional support for its relevance as a quantitative measure for land-use and mobility patterns in the Fennoscandian Mesolithic.  

By drawing on approaches that have been used for analogous treatment of radiocarbon dates, the final paper of the thesis develops a method for assessing the summed probability of multiple shoreline dates with the aim of mapping the frequency distribution of shoreline-dated sites over time. This was then compared to an analysis of the summed probability of radiocarbon dates from within the same area, as both measures have previously been suggested to be related to demographic dynamics. The development of the two proxies diverged, with the frequency of shoreline dated sites following some process of decline through the period and the frequency of radiocarbon dates commencing later than the shoreline dates, and reaching a stable plateau after an initial period of growth. Consequently, while it seems reasonable that both measures hold some demographic signal, the divergence between them would mean that some effects are confounding this relationship. In the paper, it was suggested that mobility patterns is an important determinant for the development in the frequency of shoreline dated sites over time. While mobility patterns were also suggested to impact the frequency of radiocarbon dates, it was proposed that this effect is not as substantial as for the shoreline dated sites. However, properly understanding this relationship will depend on directly assessing the influence of a range of potentially confounding effects.


\tableofcontents

\clearpage
\addcontentsline{toc}{chapter}{\listfigurename}
\listoffigures

\clearpage
\addcontentsline{toc}{chapter}{\listtablename}
\listoftables
\clearpage
\pagestyle{plain}

\pagestyle{fancy}
\fancyhead{} % Clear all header fields
\fancyfoot{} % Clear all footer fields
\fancyhead[LE,RO]{\thepage} % Page number on the left for even pages, right for odd pages
\fancyhead[RE]{\textit{Chapter \thechapter. \leftmark}} % Chapter title on the right for even pages
\fancyhead[LO]{\textit{\rightmark}} % Section title on the left for odd pages
\renewcommand{\chaptermark}[1]{\markboth{#1}{}} % Customize chapter mark
\renewcommand{\sectionmark}[1]{\markright{#1}} % Customize section mark
\renewcommand{\headrulewidth}{0.001pt}
\clearpage{\pagestyle{empty}\cleardoublepage}